A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
more sect 2

Third, \emph{it must be possible to express cyclic }(\emph{$\omega$-regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic
values internally. For example,
the term $\const{SCons}(1,\: \mu s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
denotes the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9, \ldots$\afterLdots.

@@

\paragraph{Special Cases.}
A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive; however, the
semantic notion is more fruitful and is the one we focus on.

@@

Degenerate Codatatypes

    Consider

      codatatype a = A(a)

    This is a finite codatatype, with a single value, which is infinite:
    $\mu a. A(a)$

    We need to detect those finite codatatypes. They are easy to
    characterize semantically.

    Lemma D.1. The only possible finite cardinality for the domain interpreting
    a (corecursive) codatatype is 1.

    Proof.
      Assume you have at least 2 different values of a codatatype.
      They differ at one point.
      * 
    Qed

    An immediate consequence by the distinctness axiom is that these types have
    necessarly one constructor, and that all arguments are of types of
    cardinality 1.

    Easy case: $k$ mutually recursive codatatypes, each equipped with one
    constructor, all of which take each other in arguments in various
    combinations, with potential other types that are known to have cardinality
    1. e.g.

      codatatype b = B b c b
             and c = C c b c

    Medium case: like above, but some of the arguments are other types not
    participating in the corecursion that are known to have cardinality 1, e.g.

      codatatype d = D unit d d unit

    where "unit" is a datatype with the single constructor "Unity : unit".

    Though case: like above, but with any other types where it's not clear
    whether it could have cardinality one or not.

      type e
      codatatype f = F e f f unit

@@

\rn{Ind}
\(
\inferrule{
  %x : \tau
  %\quad
  %\tau \in \Codata
  v_{\ec{x}} \in FV( \Val )
  \quad
  C( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  C \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. C( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\)

@@

As an example,
say that $\Fc$ contains equivalence classes $\ec{x}$ and $\ec{y}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$, where $z$ and $w$ are variables of integer type.
A possible application of the above rules, showing the state of $\Val$ after each application, is as follows:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\hline
\text{Rule} & \multicolumn{4}{c}{\Val} \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{z}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( \const{0}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{w}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( \const{1}, \mu v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( \const{0}, \mu v_{\ec{x}}. \const{C}( \const{1}, v_{\ec{y}} ) ), & \ec{w} \mapsto \const{1} & \} \\
\end{array}\]

>>>>

Third, \emph{it must be possible to express cyclic }(\emph{$\omega$-regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic
values internally. For example,
the term $\const{SCons}(1,\: \mu\vthinspace s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
denotes the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9, \ldots$\afterLdots.

@@

\paragraph{Special Cases.}
A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive; however, the
semantic notion is more fruitful and is the one that interests us.

@@

Finally, some codatatypes are so degenerate that they have infinite values
even though they are finite. The simplest example is
\[\keyw{datatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: $\mu\vthinspace a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{unit},\: \ty{c},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$. We call such types \emph{corecursive singletons}.
For the decision procedure, it will be important to detect those finite
codatatypes.

% In other words, all finite corecursive codatatypes are corecursive singletons.

\begin{lemma}
If the domain interpreting a corecursive codatatype is finite, it must be $1$.
\end{lemma}

\begin{proof}
TODO.
\qed
\end{proof}

An immediate consequence by the distinctness axiom is that these types have
necessarly one constructor and that all arguments have types of cardinality 1.
%
Sometimes, a type might be a corecursive singleton only in some models. For
example, if \ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be
singleton types precisely in the models that interpret \ty{unit} as a
singleton.

@@

\rn{Ind}
\(
\inferrule{
  %x : \tau
  %\quad
  %\tau \in \Codata
  v_{\ec{x}} \in FV( \Val )
  \quad
  C( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  C \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu\vthinspace v_{\ec{x}}. C( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\)

@@

As an example,
say that $\Fc$ contains equivalence classes $\ec{x}$ and $\ec{y}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$, where $z$ and $w$ are variables of integer type.
A possible application of the above rules, showing the state of $\Val$ after each application, is as follows:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\hline
\text{Rule} & \multicolumn{4}{c}{\Val} \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu\vthinspace v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{y}
& \{ & \ec{x} \mapsto \mu\vthinspace v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu\vthinspace v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu\vthinspace v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu\vthinspace v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu\vthinspace v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu\vthinspace v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu\vthinspace v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{z}
& \{ & \ec{x} \mapsto \mu\vthinspace v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu\vthinspace v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu\vthinspace v_{\ec{y}}. \const{C}( \const{0}, \mu\vthinspace v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{w}
& \{ & \ec{x} \mapsto \mu\vthinspace v_{\ec{x}}. \const{C}( \const{1}, \mu\vthinspace v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu\vthinspace v_{\ec{y}}. \const{C}( \const{0}, \mu\vthinspace v_{\ec{x}}. \const{C}( \const{1}, v_{\ec{y}} ) ), & \ec{w} \mapsto \const{1} & \} \\
\end{array}\]
